---
title: "GLMs, GLMMs, and Choosing a Distribution"
author: "P. Alexander Burnham"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    theme: cosmo
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

# Overview

This lecture uses the **Bumble Bee Disease Seasonal Variation dataset** and is organized into five core topics:

1.  **Model structure for the four main `y ~ x` data-type combinations**
2.  **Testing for significance**
3.  **Distributions and log-transformation**
4.  **Random effects and nesting**
5.  **Interaction effects**

The response variable should usually drive the choice of model family. In this dataset we can use:

-   **BQCV, DWV and Nosema** prevalence as a binary response
-   **BQCV, DWV and Nosema load** as a continuous response,
-   **Bee Species, Host Plant, and Bee Cast** as categorical predictors
-   and **Sampling Event** as a time-like predictor.
-   **Site Code** as random effects

# Load packages and import data

```{r}
library(tidyverse)
library(lubridate)
library(lme4)
library(car)
```

```{r}
bee_dat <- read_csv("data/Burnham_field_data_pathogens_wide.csv")

bee_dat <- bee_dat %>%
  mutate(
    sampling_date = mdy(sampling_date),
    site_code = factor(site_code),
    field_id = factor(field_id),
    bee_caste = factor(bee_caste),
    bombus_spp = factor(bombus_spp),
    host_plant = factor(host_plant),
    sampling_event = factor(sampling_event),
    sampling_event_num = as.numeric(as.character(sampling_event)),
    log10_BQCV_load = log10(BQCV_pathogen_load + 1),
    log10_DWV_load = log10(DWV_pathogen_load + 1),
    log10_Nosema_load = log10(Nosema_pathogen_load + 1)
  )

glimpse(bee_dat)
```

# 1. Model structure for the four main variable type combinations

A useful starting point is the standard `y ~ x` matrix.

| Response `y` | Predictor `x` | Typical figure | Common model |
|------------------|------------------|------------------|------------------|
| Continuous | Continuous | scatterplot | `lm()` / Gaussian GLM |
| Continuous | Categorical | boxplot | `lm()` / ANOVA-style model |
| Categorical | Continuous | logistic curve | `glm(..., family = binomial)` |
| Categorical | Categorical | grouped proportions | `glm(..., family = binomial)` |

## Continuous y \~ Continuous x

Use **DWV load** as the response and **BQCV** as the predictor.

```{r}

# filter for only pos:
df_filtered <- bee_dat[bee_dat$log10_DWV_load > 0 & bee_dat$log10_BQCV_load > 0, ]

qplot(
  x = log10_BQCV_load,
  y = log10_DWV_load,
  data = df_filtered) +
geom_smooth(method = "lm", se = TRUE)
```

```{r}
m_cont_cont <- lm(log10_DWV_load ~ log10_BQCV_load, data = df_filtered)
Anova(m_cont_cont)
```

## Continuous y \~ Categorical x

Use **BQCV load** as the response and **bee spp** as the categorical predictor.

```{r}
qplot(
  x = bombus_spp,
  y = log10_BQCV_load,
  data = df_filtered,
  geom = "boxplot")
```

```{r}
m_cont_cat <- lm(log10_BQCV_load ~ bombus_spp, data = df_filtered)
Anova(m_cont_cat)
```

## Categorical y \~ Continuous x

Use **BQCV presence/absence** as the response and **sampling_event** as a numeric predictor.

```{r}
qplot(
  x = log10_BQCV_load,
  y = DWV_pathogen_binary,
  data = bee_dat) +
  geom_smooth(
    method = "glm",
    method.args = list(family = "binomial"))
```

```{r}
m_cat_cont <- glm(
  DWV_pathogen_binary ~ log10_BQCV_load,
  data = bee_dat,
  family = binomial(link="logit")
)
Anova(m_cat_cont)
```

## Categorical y \~ Categorical x

Use **DWV presence/absence** as the response and **bee_caste** as the predictor.

```{r}

# summarize
sm <- bee_dat %>% 
  group_by(bee_caste) %>%
  dplyr::summarise(
    mean = mean(DWV_pathogen_binary, na.rm=T), 
  ) 

qplot(
  x = bee_caste,
  y = mean,
  data = sm
)
```

```{r}
m_cat_cat <- glm(
  DWV_pathogen_binary ~ bee_caste,
  data = bee_dat,
  family = binomial(link="logit")
)
summary(m_cat_cat)
```

# 2. Testing for significance

There is more than one way to test whether predictors improve a model.

## Let's Create a couple of new models and test for significance:

```{r}

# a model using a binomial dist.
bin_mod <- glm(data = bee_dat, DWV_pathogen_binary ~ bombus_spp * sampling_event + host_plant, family = binomial(link="logit"))

# a gaussian model:
gaus_mod <- lm(data = bee_dat, log10_Nosema_load ~ sampling_event * host_plant)


```

## Coefficient tests from the model summary

```{r}
summary(bin_mod)
summary(gaus_mod)
```

These tests ask whether individual coefficients differ from zero, given the current model structure.

## Type II or Type III tests with `car`

```{r}
Anova(bin_mod, type = 2)
Anova(gaus_mod, type = 2)
```

These are often useful when predictors have multiple levels or when you want term-level tests rather than coefficient-level tests.

-   **Type 1 (Sequential/Hierarchical):** Evaluates predictors in a specific order, adding them one by one. The order matters, as each variable is tested based on what is already in the model, commonly used for testing hierarchical models.

-   **Type 2 (Marginal/Hierarchical):** Tests each effect after all other *appropriate* effects are in the model (e.g., testing main effects with interaction terms included, but not testing higher-order interactions simultaneously). It does not assume a specific order.

-   **Type 3 (Partial/Marginal):** Evaluates each effect after adjusting for all other effects in the model, regardless of order. It is generally the default in many software packages for ANOVA and regression, testing if a variable adds value to the model over all others.

## Likelihood ratio test: full vs null model

```{r}
m_dwv_null <- lm(data = df_filtered, log10_DWV_load ~ 1)
m_dwv_full <- lm(data = df_filtered, log10_DWV_load ~ sampling_event + host_plant)

anova(m_dwv_null, m_dwv_full, test = "LRT")
```

This asks whether the predictors improve fit compared with a model containing only an intercept.

## Likelihood ratio test: full vs reduced model

```{r}
m_dwv_reduced <- lm(data = df_filtered, log10_DWV_load ~ sampling_event)

anova(m_dwv_reduced, m_dwv_full, test = "LRT")

```

This asks whether the additional term improves the model beyond what is already included.

## Term deletion with `drop1()`

```{r}
drop1(m_dwv_full, test = "Chisq")
```

This is a convenient way to evaluate each term within the same model.

# 3. Distributions and log transformation

The response variable should usually determine the distribution.

## First, look at the response distributions

### Binary response: BQCV presence

```{r}
qplot(
  x = factor(BQCV_pathogen_binary),
  data = bee_dat,
  geom = "bar")
```

A 0/1 response is naturally modeled with a **binomial GLM or GLMM**.

### Continuous raw response: DWV load

```{r}
qplot(
  x = DWV_pathogen_load,
  data = df_filtered,
  geom = "histogram",
  bins = 30)
```

Raw pathogen loads are often strongly right-skewed.

### Log-transformed continuous response: DWV load

```{r}
qplot(
  x = log10_DWV_load,
  data = df_filtered,
  geom = "histogram",
  bins = 30)
```

The log transform often makes the distribution more symmetric and easier to model with a Gaussian response.

### Continuous raw response: Nosema load

```{r}
qplot(
  x = Nosema_pathogen_load,
  data = bee_dat,
  geom = "histogram",
  bins = 30)
```

## Distribution guide

-   Use **Gaussian** models for continuous responses that are roughly symmetric, often after transformation.
-   Use **binomial** models for 0/1 responses such as pathogen presence.
-   Use **Gamma** models for positive continuous responses when you want to model the raw scale directly.
-   Use **Poisson** or **negative binomial** for count responses, which are not the main response type in this dataset.

## Why log-transform?

A log transform is useful when:

-   the response is continuous and positive,
-   the raw data are strongly right-skewed,
-   variance increases with the mean,
-   and interpretation on a multiplicative scale makes sense.

```{r}
m_dwv_log <- lm(log10_DWV_load ~ sampling_event + bee_caste, data = bee_dat)
summary(m_dwv_log)
```

Do **not** log-transform binary responses. Also, logging is not mandatory if a different GLM family is more natural for the response.

# 4. Random effects and nesting

GLMMs extend GLMs by adding **random effects**, which account for grouped or non-independent observations.

In field biology, non-independence often arises because samples are collected within the same **site**, **field**, or other grouping level.

## Why random effects?

Use random effects when:

-   observations are clustered,
-   you expect groups to differ in baseline response,
-   and the group levels are better treated as a sample from a broader population rather than as fixed categories of direct interest.

## Random intercept example

Here is a binomial GLMM for BQCV presence with a random intercept for site.

```{r}
g_bqcv_site <- lmer(
  log10_BQCV_load ~ bombus_spp + sampling_event + (1 | site_code),
  data = df_filtered)

Anova(g_bqcv_site)
```

## Nested random effects

If fields are nested within sites, we can write that as `field_id` within `site_code`.

```{r}
g_bqcv_site <- lmer(
  log10_BQCV_load ~ bombus_spp + (1 | site_code/sampling_event),
  data = df_filtered)

Anova(g_bqcv_site)
```

This expands to random intercepts for site ID and for sampling event nested within site.

## Gamma mixed model example

We can also use a mixed model for a different response error dist.

```{r}

# make pos only nosema
nosPos <- bee_dat[bee_dat$Nosema_pathogen_load > 0,]

# gamma
nos_gamma <- glmer(
  Nosema_pathogen_load ~ site_code + bombus_spp + (1 | sampling_event),
  data = nosPos, family = Gamma)
Anova(nos_gamma)

# log
nos_log <- lmer(
  log10_Nosema_load ~ site_code + bombus_spp + (1 | sampling_event),
  data = nosPos)
Anova(nos_log)
```

In practice, random effects are justified both statistically and biologically. If repeated samples come from shared sites or fields, independence is violated, and mixed models are often the better choice.

# 

# 5. Interactions

Interactions ask whether the effect of one predictor depends on the level of another predictor.

Because **sampling_event** is time-like, it is a useful interaction term in this dataset.

## Example 1: interaction in a Gaussian model

Does the change in DWV load across sampling event differ by caste?

```{r}
qplot(
  x = sampling_event_num,
  y = log10_DWV_load,
  color = bee_caste,
  data = bee_dat)
```

```{r}
m_dwv_interaction <- lmer(log10_DWV_load ~ sampling_event * bee_caste + (1 | site_code), data = bee_dat)
summary(m_dwv_interaction)
Anova(m_dwv_interaction, type = 2)
```

## Example 2: interaction in a binomial model

Does the probability of BQCV infection across sampling event differ by caste?

```{r}
m_bqcv_interaction <- glm(
  BQCV_pathogen_binary ~ sampling_event * log10_DWV_load,
  data = bee_dat,
  family = binomial(link="logit")
)
summary(m_bqcv_interaction)
Anova(m_bqcv_interaction, type = 2)
```

You can also compare the interaction model to the additive model:

```{r}
m_bqcv_additive <- glm(
  BQCV_pathogen_binary ~ sampling_event + log10_DWV_load,
  data = bee_dat,
  family = binomial(link="logit")
)

summary(m_bqcv_additive) # higher aic
Anova(m_bqcv_additive, type = 2)

anova(m_bqcv_additive, m_bqcv_interaction, test = "Chisq")
```

# Closing summary

-   Start by identifying whether the response is **continuous** or **categorical**.
-   Use the `y ~ x` matrix to choose a basic model structure.
-   Inspect the **distribution of the response** before choosing a family.
-   Use **log transformation** for right-skewed continuous responses when appropriate.
-   Use **interactions** when the effect of one predictor may depend on another.
-   Use **random effects** when observations are grouped or nested.
-   Test significance using several complementary tools: `summary()`, `car::Anova()`, likelihood ratio tests, and reduced-model comparisons.