---
title: "Statistical Machine Learning"
subtitle: "PCA and Random Forest in R"
author: P. Alexander Burnham
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

# Agenda
- framing — why PCA vs random forest
- Coding up a PCA
- Coding up a random forest model

# Framing

## PCA vs. Random Forest

- **PCA is unsupervised**: no response variable; goal is *structure + variance reduction*.
- **Random forest is supervised**: response variable exists; goal is *prediction/classification + generalization*.

## What you should pay attention to in each workflow

- **PCA**: scaling, variance explained, scores vs loadings, “how many axises”, and whether structure is meaningful.
- **Random forest**: train/test split (or CV), OOB (out-of-bag error), confusion matrix, variable importance.

# Setup up our environment:

```{r}
# For reproducible results
set.seed(1)

# Packages used today
library(ggplot2)
library(psych)        # Bartlett's test
library(randomForest) # Random forest

# Data
data(iris)
```

Quick peek:

```{r}
head(iris)
str(iris)
```



# Part I — PCA (unsupervised)

## Step 1: choose numeric columns and (usually) scale

PCA uses **numeric** variables. Factors like `Species` are great for *coloring plots* later, but not for computing PCs.

```{r}
iris_num <- iris[, 1:4]
iris_species <- iris$Species

round(cor(iris_num), 2)
```

## Step 2: check whether PCA is “worth it” (Bartlett’s test)

Bartlett’s test asks whether the correlation matrix differs from the identity matrix.

```{r}
bart <- cortest.bartlett(cor(iris_num), n = nrow(iris_num))
bart
```

Expected output (trimmed):

```
$chisq
[1] ...

$p.value
[1] ...
...
```

## Step 3: fit PCA with `prcomp()`

We scale because sepal/petal variables are on different ranges.

```{r}
pca <- prcomp(iris_num, center = TRUE, scale. = TRUE)
summary(pca)
```

Expected output (trimmed):

```
Importance of components:
                          PC1    PC2    PC3    PC4
Standard deviation     ...    ...    ...    ...
Proportion of Variance ...    ...    ...    ...
Cumulative Proportion  ...    ...    ...    ...
```

## Step 4: variance explained (small table)

```{r}
eig <- pca$sdev^2
pve <- eig / sum(eig)

pca_var_table <- data.frame(
  PC = paste0("PC", 1:length(eig)),
  Eigenvalue = round(eig, 3),
  PVE = round(pve, 3),
  CumPVE = round(cumsum(pve), 3)
)
pca_var_table
```

## Step 5: scree plot

```{r}
plot(eig, type = "b", pch = 19,
     xlab = "Principal component",
     ylab = "Eigenvalue",
     main = "Scree plot (iris PCA)")


# test
broken_stick <- function(p) sapply(1:p, function(k) sum(1/(k:p)) / p)
bs <- broken_stick(ncol(iris_num))

retain <- data.frame(
  PC = paste0("PC", 1:length(pve)),
  ObservedPVE = round(pve, 3),
  BrokenStick = round(bs, 3),
  Keep = pve > bs
)
retain
```

## Step 6: interpret scores and loadings

- **Scores**: where each observation lands in PC space.
- **Loadings**: how strongly each original variable contributes to each PC.

```{r}
head(pca$x)
pca$rotation
```



## Step 7: plot PC1 vs PC2 (color by species)

```{r}
scores <- as.data.frame(pca$x)
scores$Species <- iris_species

plt <- ggplot(scores, aes(PC1, PC2, color = Species)) +
  geom_point(size = 2.6, alpha = 0.85) +
  theme_minimal() +
  labs(title = "PCA on iris", subtitle = "PCA is unsupervised; species used only for coloring")

plt + stat_ellipse() # 95% CI


```


## Optional: test group separation *after* PCA

PCA itself is unsupervised. But after PCA, you can test whether known groups differ in the reduced space.

```{r}
man <- manova(cbind(PC1, PC2) ~ Species, data = scores)
summary(man, test = "Pillai")
```

# Part II — Random forest (supervised)

## Step 1: define response and predictors

Here the response is **Species** (categorical), and predictors are the four measurements.

## Step 2: train/test split

```{r}
set.seed(42)
id_train <- sample(seq_len(nrow(iris)), size = 0.7 * nrow(iris))
train <- iris[id_train, ]
test  <- iris[-id_train, ]
```

## Step 3: fit a random forest classifier

```{r}
set.seed(123)
rf <- randomForest(
  Species ~ ., data = train,
  ntree = 500,
  mtry = 2,
  importance = TRUE
)
rf
```



## Step 4: evaluate on the test set

```{r}
pred <- predict(rf, newdata = test)
conf <- table(Observed = test$Species, Predicted = pred)
conf

acc <- mean(pred == test$Species)
acc
```

## Step 5: OOB error curve

OOB error is an internal “almost cross-validation” estimate from bootstrap sampling.

```{r}
plot(rf, main = "Random forest OOB error vs number of trees")
```

## Step 6: variable importance

```{r}
importance(rf)
varImpPlot(rf)
```


## Step 7 (quick): compare to a single decision tree intuition

Talking point (no extra code needed):

- A single tree can be unstable.
- A forest averages many de-correlated trees → typically improves generalization.

